\documentclass[a4page]{article}
% \usepackage{biblatex}
\usepackage[a4paper,
            bindingoffset=0.2in,
            left=1in,
            right=1in,
            top=1in,
            bottom=1in,
            footskip=.25in]{geometry}
\usepackage{amsmath, amsfonts, amsthm , mathtools , bbold, float}

\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage[citecolor=blue,linkcolor=blue,urlcolor=blue,colorlinks]{hyperref}
\usepackage[nameinlink,capitalize,noabbrev]{cleveref}
\usepackage{quiver}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{bbm}
\usepackage{breakurl}
\usepackage{bbding}
\hypersetup{pdfborder={0 0 100}}
\usepackage{lipsum}
%\usepackage{biblatex}
%\bibliographystyle{unsrt}
\usepackage[backend=biber,style=alphabetic,sorting=nty,giveninits]{biblatex} % Bibliography management
\addbibresource{refs.bib}

\usepackage{textcomp}
\usepackage{xspace}
\usepackage{url}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{pifont}
\usetikzlibrary{arrows,matrix,decorations.pathmorphing,
  decorations.markings, calc, backgrounds}
\usepackage{mathpartir}
\usepackage{microtype}

\DisableLigatures[-]{family=tt*}

\input{macros}

\title{Hurewicz and Brouwer}

\author{
  Axel Ljungström\\
  \footnotesize{Stockholm University, Sweden}\\
  \footnotesize{\texttt{axel.ljungstrom@math.su.se}}
  \and
  Loïc Pujet\\
  \footnotesize{Stockholm University, Sweden}\\
  \footnotesize{\texttt{loic@pujet.fr}}
}
\date{}

\begin{document}

\maketitle
\vspace{-.5cm}
%
Witold Hurewicz is one of the most important names in algebraic topology.
Most famously, he is credited with the definition of higher homotopy groups, and with the
\emph{Hurewicz theorem} which establishes a relation between homology groups and these
newly defined homotopy groups.
%
Interestingly, Hurewicz worked from 1927 to 1936 as an assistant of L.E.J. Brouwer,
who is best known as the father of intuitionism.
%
However, it seems that Hurewicz did not inherit his mentor's interest for constructive proofs.
%
In this abstract, we will do our best to posthumously reconcile these two mathematicians
by giving a constructive proof of the Hurewicz theorem in Homotopy Type Theory.

We should note that Christensen and Scoccola have already proved a version of the Hurewicz
theorem in HoTT~\cite{christensen2020hurewicz}.
They work with Eilenberg-MacLane homology, and their proof boils down to connectivity
properties of the smash product.
%
On the other hand, the proof that we give here uses the more elementary notion of
\emph{cellular homology}, and the core of our argument is that \( n \)-connected spaces may be
approximated by \emph{CW complexes} with no nontrivial cells in dimension~\( \le n \).
%
We use the definition of CW complexes that was introduced by Buchholtz and Favonia~\cite{BuchholtzFavonia18}:

\begin{figure}[H]
\begin{minipage}{0.68 \linewidth}
\begin{definition}
  A \textbf{CW structure} is an infinite sequence of types and maps
  $(X_{-1} \xrightarrow{\incl{-1}} X_0 \xrightarrow{\incl{0}} X_1 \xrightarrow{\incl{1}} \dots )$
  equipped with a function $c : \bN \to \bN$ and a set of attaching maps $\alpha_i : S^i \times \fin{c_{i+1}} \to X_i$ for
\end{definition}
\end{minipage}
\vspace{-.1cm}
\hspace{.1cm}
\begin{minipage}{0.32 \linewidth}
  %
\begin{tikzcd}[ampersand replacement=\& , column sep = small , row sep = small]
	{S^{i} \times \fin{c_{i+1}}} \&\& \fin{c_{i+1}} \\
	{X_i} \&\& {X_{i+1}}
	\arrow["{\mathsf{snd}}", from=1-1, to=1-3]
	\arrow["{\alpha_{i}}"', from=1-1, to=2-1]
	\arrow[from=2-1, to=2-3]
	\arrow[from=1-3, to=2-3]
	\arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-1]
\end{tikzcd}
\end{minipage}
\emph{$i \geq -1$ s.t. $X_{-1}$ is empty and the square on the right is a (homotopy) pushout.
  A type $A$ is said to be a \textbf{CW complex} if there merely exists some
  CW structure $X_\ast$ s.t.\ $A$ is equivalent to the sequential colimit
  of $X_\ast$, i.e.\ $A \simeq X_\infty$.}
\end{figure}
%
Informally, a CW complex is a type that can be built by an iterative process of attaching
cells: we start with a finite number of 0-dimensional points, then we add a finite
amount of 1-dimensional edges, then we add a finite amount of 2-dimensional discs, and so on.
%
These explicit descriptions are quite convenient for defining properties and operations by
induction on dimensions. Most notably, Buchholtz and Favonia showed how to define the
\emph{cellular homology groups} of any CW complex.

\begin{proposition}
  Given any integer \( n \ge 0 \), one can define a functor \( \widetilde{H}_n \) from the
  category of CW complexes to the category of abelian groups.
  Moreover, the resulting family of functors satisfy the Eilenberg-Steenrod axioms for reduced homology.
\end{proposition}

Our goal is to prove the \emph{cellular Hurewicz theorem}, which informally states that if a CW complex is
\( (n - 1) \)\nobreakdash-connected, then its cellular homology groups coincide with its homotopy groups up to
dimension \( n \) (up to abelianisation in the case \( n = 1 \)).
%
The classical proof of this theorem takes an arbitrary \( (n - 1) \)-connected CW complex, and replaces
its CW structure with an alternative one with no nontrivial cells in dimension~\( < n \).
%
This is done by defining the new set of \( n \)-cells to be the \( n \)th
homotopy group of the space, from which the Hurewicz theorem will follow.
%
Unfortunately, this approach will not work with our definition of CW complexes which only allows
a finite number of cells in every dimension, since finite CW complexes tend to have infinite
homotopy groups (think of the \( n \)-sphere for instance).
%
Thankfully, it turns out that a slightly different approximation theorem is provable without leaving
the world of finite CW complexes.

\begin{definition}
  We say that a CW structure $X_\ast$ is \textbf{Hurewicz $n$-connected} if $c^X_0 = 1$ and $c^X_i = 0$ for $0<i\le n$.
  We use the same terminology for CW complexes which merely have a Hurewicz $n$-connected CW structure.
\end{definition}

\begin{theorem}\label{cor:hurewicz-con}
  A CW complex is $n$-connected if and only if it is Hurewicz $n$-connected.
\end{theorem}

\begin{proof}
A formal proof is available at \url{https://github.com/agda/cubical/blob/master/Cubical/CW/Connected.agda}.
\end{proof}

With this approximation theorem, we are in position to compute the homology and the homotopy
of \( (n - 1) \)-connected CW complexes.
%
First, say that a type is a \( n \)-dimensional \emph{sphere bouquet} if it can be
written as \( \bouquetGen{A}{n} \) for some finite set \( A \).
%
Then, \cref{cor:hurewicz-con} implies that for any \( (n - 1) \)-connected CW complex \( X_\ast \), the
\( (n+1) \)-skeleton \( X_{n+1} \) is equivalent to the homotopy cofibre of a pointed map of
\( n \)-dimensional sphere bouquets:
%
\[
X_{n+1} \simeq \mathsf{cofib}\Big( \bouquetGen{A}{n} \xrightarrow{\ f\ }_\star \bouquetGen{B}{n} \Big).
\]
%
Now, remark that \( \hredcw{n}(X_\ast) = \hredcw{n}(X_{n+1})\) for connectivity reasons.
%
We can compute this group using the exactness property of cellular homology:
consider the sequence
%
\[
\bouquetGenL{A}{n} \xrightarrow{f} \bouquetGenL{B}{n} \xrightarrow{\cfcod} C_f
\xrightarrow{\cfcod} C_{(\cfcod : \bouquetGenL{B}{n} \to C_f)} \simeq \bouquetGenL{A}{n+1}
\]
%
where \( C_f \) denotes the homotopy cofibre of \( f \), \( \cfcod \) denotes the embedding
of the codomain into the homotopy cofibre, and the final equivalence is a standard characterisation
of $X_{n+1}/X_{n}$ for CW structures~\cite{BuchholtzFavonia18}.
%
This is a cofibre sequence, and so the following sequence is exact:
%
\[
\!\!\hredstr{n}\Big(\bouquetGen{A}{n}\Big) \xrightarrow{f_*} \hredstr{n}\Big(\bouquetGen{B}{n}\Big) \xrightarrow{\cfcod_*} \hredstr{n}{(C_f)} \to 0
\]
%
where the final $0$ comes from that fact that $\hredstr{n}$ vanishes on $\bouquetGen{A}{n+1}$.
We can compute the first two homology groups using additivity, and thus we see that
$\hredstr{n}(C_f) \cong \bZ[B]/\bZ[A]$.
%
Similarly, the group $\pi_n(X_\ast)$ is equal to \( \pi_n(X_{n+1}) \), which is computed
by the following lemma.

\begin{proposition}\label{prop:hurewicz-seq}
  For any $f : \bouquetGen{A}{n} \to \bouquetGen{B}{n}$ where $n\geq 1$ and $A$ and $B$ are finite
  types, there is an exact sequence
  \begin{center}
    \( \pi_n\left(\bouquetGen{A}{n}\right) \xrightarrow{f_*} \pi_n\left(\bouquetGen{B}{n}\right)
       \dhxrightarrow{\cfcod_*} \pi_n(C_f). \)
  \end{center}
\end{proposition}

\begin{proof}
  This follows from the Seifert-Van Kampen theorem \cite[Example 8.7.17]{HoTT13} in the case
  \( n = 1 \), and from the Blakers-Massey theorem~\cite{FavoniaFinster+16} in the case \( n > 1 \).
\end{proof}

The case \( n = 1 \) is a bit special: the fundamental group of a bouquet of circles is a
free group rather than a free abelian group. Therefore, in order to match what happens with homology,
we introduce the \emph{abelian homotopy group} functor \( \ab{n} \) which is simply defined as the
abelianisation of \( \pi_n \).
%
With this minor correction, we obtain that $\ab{n}(C_f) \cong \bZ[B]/\bZ[A]$.
%
Thus, we have proved that the \( n \)th homology group coincides with the abelianisation of the
\( n \)th homotopy group for a \( (n - 1) \)-connected CW complex.
%
The Hurewicz theorem actually goes a step further that this, and gives an explicit description
for the isomorphism between these two groups.

\begin{theorem}
  Let $X$ be a CW complex.
  Define the \textbf{Hurewicz homomorphism} $\hur : \pi_n(X) \to \hredcw{n}(X)$ on canonical
  elements $f : \sphere{n} \to_\star X$ by letting $\hur(\trunc{f}) : \hredcw{n}(X)$ be the image
  of $1$ under the composition
  \( \bZ\ \raisemath{-1pt}{\xrightarrow{\sim}}\ \hredcw{n}(\sphere{n})
  \ \raisemath{-1pt}{\xrightarrow{f_*}}\ \hredcw{n}{(X)} \).
\end{theorem}

\begin{theorem}
  The Hurewicz homomorphism $\hur : \ab{n}(X) \to \hredcw{n}(X)$ is an isomorphism for any
  $(n-1)$-connected CW complex $X$.
\end{theorem}

\begin{proof}
  A formal proof is available at \url{https://github.com/loic-p/cellular/blob/main/Hurewicz/Map.agda}
  (\texttt{HurewiczTheorem}).
\end{proof}

In conclusion, we have a fully mechanised proof of the Hurewicz theorem for cellular homology.
%
This development was motivated by the recent proof of the Serre Finiteness theorem by Barton and
Campion~\cite{SerreFiniteness}, which relies on homology computations and the Hurewicz theorem.
%
We hope that our formal development will be helpful to the ongoing formalisation of the Serre
Finiteness theorem (by, in particular, Milner~\cite{MilnerTalk}).

\printbibliography


\end{document}
